Whole-Body Manipulation Planning for Humanoid Robots

### 2012-06-15 paper.tex
Add Nishiwaki footstep reference
Add a picture to the small space controllability proof. Minor corrections

The problem of inverse kinematics for a humanoid robot, or any articulated
structure, is to compute a joint motion to achieve an end-effector pose. As the
robots we deal with are redundant, it is natural to take advantage of
this redundancy by specifying multiple tasks, potentially with
different priorities. This problem has been widely studied in robotics
planning and control literature, and many jacobian-based solutions have been
proposed, among which 
\cite{nakamura1986iks}, \cite{siciliano1991gfm},
\cite{baerlocher1998tpf} and \cite{khatib2004wbd}.
Obstacle avoidance can be taken into account with similar methods. To
do so, one has to include the obstacles as other task constraints to
be satisfied. A recent  contribution on this subject is \cite{kanehiro2008lca}.
These methods are prone to fall into local minima, global motion 
planning is needed to overcome this limitation.

@@

Simplifying  the environment  consists in  considering obstacles  at a
footstep   level.   \cite{chestnutt2005footstep,kuffner2005motion}
use   an  
A$^{*}$   algorithm  to   find
collision-free   footsteps.    In   \cite{perrinbiped},   the   authors   compute
collision-free motions  for the legs by using  an RRT$^{*}$ algorithm.

@@

The main contribution of this work is a  whole-body motion planner for humanoid robots
that computes collision-free walking trajectories, based on
exact models of both robot and environment. It is used to solve manipulation tasks
that may require walking. The first stage of our algorithm uses a
constrained sampling-based motion 
planner and  computes a collision-free path  for a statically stable robot on the
ground. Another  contribution of this  paper is the formal  proof that
this first path can always be  approximated by a dynamically balanced, collision-free
walking  trajectory in a second stage. We have implemented this well-grounded 
method and  used it on a model of  HRP-2 robot.  The results have been  
validated on a real platform.

@@

This section presents an algorithm for constrained motion planning. 
The problem solved here
differs from classic configuration approaches in two ways:
\begin{enumerate}
\item the set of valid configurations is defined implicitly, as the ones satisfying a given 
  set of inverse kinematic tasks;
\item the goal is also defined implicitely, by another set of inverse kinematic tasks.
\end{enumerate}
In this presentation, we use a prioritized inverse kinematic solver, or local planner,
 as a black box. We will not describe the details of its implementation. They are presented in 
\cite{yoshida2006tds}, and are out of scope of the present work.

@@

We present in Fig.~\ref{fig:wb-shelves} an illustration of 
the use of randomized motion planning 
on complex manipulation problems. The humanoid robot HRP-2, faces shelves. It has to: (i) grasp
a ball lying on a shelf, (ii) put it on a higher shelf, (iii) come back to a natural
rest configuration. The statically stable manifold on which the planning takes place is
defined by tasks constraining the robot feet and CoM. The goal of the first problem
is defined by a hand configuration task (position and orientation), and a gaze task
(the robot has to see the ball). The goal of the second problem is defined by a 
different hand configuration and the same gaze task,
and finally, the goal of the third problem is the rest configuration.
Extension 1 presents a video of this motion.

@@

The motion presented in Fig.~\ref{fig:wb-shelves} looks very complicated.
The reasons are that the environment is complex and HRP-2 kinematic chain is highly constrained.
The motion has been optimized based on these constraints. Note that the
accessibility space of this robot is not wide enough to accomplish this task further 
away from the shelves, which could have ease the motion. One possibility to extend the
robot accessibility space is to automatically change the support polygon to solve the task.
This is an ongoing research field, \cite{OussamaKanoun06102010} proposes some 
local solutions. This has not been integrated in our global planning algorithm yet.

@@

A  robotic system  is  controllable  if  for any two  configurations
$q_1$ and $q_2$,  there exists  a
trajectory  going  from  $q_1$ to  $q_2$.  It  is  
\textit{small-space  controllable} if for  any configuration  $q$, 
for any $\epsilon >0$, there
exist $\eta >0$ such that all the configurations contained in the ball of center
$q$ and radius $\eta$ are reachable by trajectories included in the
ball of center $q$ and radius $\epsilon$.

@@

Fig. \ref{fig:stc} shows an example of collision-free 
path approximation by admissible collision-free sub-trajectories. The fact 
that this algorithm  converges is guaranteed by the small-space
controllability property.

@@

  \begin{minipage}{3cm}
    \input{stc1.tex}
  \end{minipage}
  \begin{minipage}{4cm}
    \input{stc2.tex}
  \end{minipage}

  \caption{Small-space controllability in motion planning. On the left,
    the local property: any configuration $q'$ at distance less than
    $\eta$ is reachable from $q$ by an admissible trajectory included in
    a ball of size $\epsilon$. On the right, a collision-free path from
    $q_1$ to $q_2$ is approximated by collision-free and admissible
    trajectories by using the local property.
  }
  \label{fig:stc}
\end{figure}

@@

\begin{proof}
  Let $\mathcal{C}$ be the configuration space of a small-space controllable robot, and 
  $\mathcal{C}_{free} \subset \mathcal{C}$ the set of collision-free configurations. We
  consider in-contact configurations as colliding, so $\mathcal{C}_{free}$ is an open set.
  Let $\tau : [0,1] \rightarrow \mathcal{C}_{free}$ be a collision-free path. For every $x \in [0,1]$,
  $\tau(x) \in \mathcal{C}_{free}$, so there exists $\epsilon_x$ such that the open ball 
  $B(\tau(x),\epsilon_x)$ of center $\tau(x)$ and radius $\epsilon_x$ is included in 
  $\mathcal{C}_{free}$. The small-space controllability property states that for every $x$,
  there exist $\eta_x$ such that every configuration $q \in B(\tau(x),\eta_x)$ is reachable 
  from $\tau(x)$ by a trajectory included in $B(\tau(x),\epsilon_x)$.

@@

\begin{theorem}
A quasi-static walking robot is not small-space controllable. A dynamic walking robot is.
\end{theorem}

\begin{proof}

The first claim is straightforward. Let the robot be in a configuration $q$ where the two
feet are separated from a positive distance. Let us assume that the CoM is at a horizontal
distance $L>0$ from the left foot (if the CoM is over the left foot, we can consider similarly 
the right foot). For any $\epsilon < L$, any valid trajectory starting from $q$, included in
the ball of center $q$ and radius $\epsilon$, is such that the CoM is never over the left foot.
Given the quasi-static walk conditions, the right foot of the robot is fixed along 
the trajectory. Thus, the set of accessible configurations from $q$ by staying inside 
$B(q,\epsilon)$ does not form a neighborhood of $q$. This shows that the robot is 
not small-space controllable.

@@

Without loss of generality, let us assume that the robot is in a configuration
in which the CoM is at the horizontal position $(0,0)$, the center of the feet are 
aligned with the $y$-axis and the horizontal distance between the CoM and either of the centers
of the feet is $L$. To achieve dynamically balanced walking, we aim at making
$p_y(t)$ oscillate  between $-L$ and $L$. To move the ZMP  under a
given foot, only  the $y$ coordinate of the CoM  is of interest. Thus,
we will keep the $x$ coordinates  of the CoM and ZMP constant equal to
$0$. We have made the hypothesis that the feet were of positive surface,
let $l>0$ be such that the length of the section of a foot along the 
$y$-axis is greater than $l$.

The idea of this proof is to use the form of Eq. (\ref{eq:walk-zmp}) to
apply a  scaling factor between  the amplitude of the  oscillations of
the CoM and of the ZMP. For $\omega >0$, if the CoM follows a trajectory given
by  $y(t) = \epsilon \sin(\omega t)$,  Eq. (\ref{eq:walk-zmp}) gives:
\[
p_y(t) =
(1+\left(\frac{\omega}{\omega_0}\right)^2)\epsilon\sin(\omega t)
\]

The
amplitude  of  the oscillations  of  $y$  is  multiplied by  a  factor
$(1+\left(\frac{\omega}{\omega_0}\right)^2)$.  If we choose  $\omega =
\omega_0 \sqrt{\frac{L}{\epsilon} -1}$,  $p_y$ oscillates between $-L$
and    $L$.   At    time   $t_l^{(n)}    =    n\frac{2\pi}{\omega}   +
\frac{\pi/2}{\omega}$, the  ZMP is located  at the center of  the left
foot,  the robot  can move  its right  foot and  at time  $t_r^{(n)} =
n\frac{2\pi}{\omega}  + \frac{3\pi/2}{\omega}$ the  ZMP is  located at
the center of the right foot, the robot can move its left foot.

@@

One can  check that $y$  is of class $C^2$ over  $\mathbb{R}_+$, and
that $\dot{f}(0) = 0$. When $t\geq T$, the robot is in the permanent
state described above  and can successively move its feet  inside small neighborhoods.
The  last point to check
is that for $t \in  [0,T]$ $p_y(t)$ stays inside the support polygon
of the robot. The calculation of the successive derivatives of $y$ gives:

@@

We have thus exhibited a continuous control scheme that allows to move any of the feet
while keeping the CoM inside an arbitrarily small neighborhood, this concludes
the proof.
\end{proof}

\subsubsection{Remarks}
\paragraph{Generalization to a complete model:} More work is needed to generalize 
the previous proof to any legged robot model. First, because we use a model where
the mass is punctual, we did not have to consider the angular-momentum of the robot
around the CoM, while it accounts for dynamic balance. Moreover, when considering
an actual kinematic and dynamic model,
the positions of feet and CoM are not controlled degrees of freedom, but inputs 
for inverse kinematic tasks. Continuity and regularity results about the robot
controller -- i.e. how much the configuration changes when the input of an inverse
kinematic task changes--, would be needed to generalize the proof. We have not
proven such properties. The  proof presented above still gave the intuition for the
algorithm presented in the following.

\paragraph{Use of ZMP preview controller:} The control strategy presented in the  
previous proof may
generate very  long trajectories, because  of the transient  states at
the beginning and end of the locomotion. In the actual implementation,
we chose  to generate  CoM motions with  a ZMP preview  controller, as
presented in \cite{kajita2003biped}.  We  have observed experimentally that the
amplitude of  CoM trajectories decreases  when the frequency  of steps
increases.

\paragraph{Speed of CoM:} The theoretical result presented in this section implies
that any collision-free path can be approximated by a sequence of admissible 
and collision-free trajectories. However, the theorem depends on a control law
that will generate trajectories with unbounded velocities for the CoM, when the input
path is close to obstacles. The humanoid robot
hardware may be a limitation to such trajectories. To prevent the generated CoM 
oscillations from being too fast, one has to require that the statically stable 
path is included
inside an $\epsilon$-radius tube of the free space, where $\epsilon$ depends on the
physical capacities of the robot.

@@

The  algorithm   that  animates  a  statically  stable   path  into  a
dynamically stable  walk trajectory has been inspired  by the previous
small-space controllability  proof. Given a statically  stable path $p$,
we  start  by placing  footsteps  corresponding  to  the nominal  walk
pattern  of the  robot. Given the footsteps, we  compute a ZMP trajectory, 
and a preview controller outputs  a corresponding CoM trajectory, using
the cart-table model.
The stack of tasks applied to
the robot is - in decreasing priority order:

@@

\item Position and orientation of  feet,

>>>>

The problem of inverse kinematics for a humanoid robot, or any articulated
structure, is to compute a joint motion to achieve an end-effector pose. As the
robots we deal with are redundant, it is natural to take advantage of
this redundancy by specifying multiple tasks, potentially with
different priorities. This problem has been widely studied in robotics
planning and control literature, and many jacobian-based solutions have been
proposed, among which 
\cite{nakamura1986iks}, \cite{siciliano1991gfm},
\cite{baerlocher1998tpf} and \cite{khatib2004wbd}.
Obstacle avoidance can be taken into account with similar methods. To
do so, one has to include the obstacles as  constraints to
satisfy, see for example \cite{kanehiro2008lca}.
These methods are prone to fall into local minima, global motion 
planning is needed to overcome this limitation.

@@

Simplifying  the environment  consists in  considering obstacles  at a
footstep   level only.   \cite{kuffner2001footstep,chestnutt2005footstep,kuffner2005motion}
use   an  A$^{*}$   algorithm  to   find
collision-free   footsteps.    In   \cite{perrinbiped},   the   authors   compute
collision-free motions  for the legs by using  an RRT$^{*}$ algorithm.

@@

The main contribution of this work is a  whole-body motion planner for humanoid robots
that computes collision-free walking trajectories, based on
exact models of both robot and environment. It is used to solve manipulation tasks
that may require walking. The first stage of our algorithm uses a
constrained sampling-based motion 
planner and  computes a collision-free  statically stable path  for a robot fixed or sliding 
on the ground. Another  contribution of this  paper is the formal  proof that
this first path can always be  approximated by a dynamically balanced, collision-free
walking  trajectory in a second stage. We have implemented this well-grounded 
method and  used it on a model of  HRP-2 robot.  The results have been  
validated on a real platform.

@@

This section presents an algorithm for constrained motion planning. 
The problem solved here
differs from classic configuration space approaches in two ways:
\begin{enumerate}
\item the set of valid configurations is defined implicitly, as the ones satisfying a given 
  set of inverse kinematic tasks;
\item the goal is also defined implicitely, by another set of inverse kinematic tasks.
\end{enumerate}
In this presentation, we use a prioritized inverse kinematic solver, or local planner,
 as a black box. We will not describe the details of its implementation. They are presented in 
\cite{yoshida2006tds}, and are out of scope of the present work.

@@

We present in Fig.~\ref{fig:wb-shelves} an illustration of 
the use of randomized motion planning 
on complex manipulation problems. The humanoid robot HRP-2 faces shelves. It has to: (i) grasp
a ball lying on a shelf, (ii) put it on a higher shelf, (iii) come back to a natural
rest configuration. The statically stable manifold on which the planning takes place is
defined by tasks constraining the robot feet and CoM. The goal of the first problem
is defined by a hand configuration task (position and orientation), and a gaze task
(the robot has to see the ball). The goal of the second problem is defined by a 
different hand configuration and the same gaze task,
and finally, the goal of the third problem is the rest configuration.
Extension 1 presents a video of this motion.

@@

The motion presented in Fig.~\ref{fig:wb-shelves} looks very complicated.
The reasons are that the environment is complex and HRP-2 kinematic chain is highly constrained.
The motion has been optimized based on these constraints. Note that the
accessibility space of this robot is not wide enough to accomplish this task further 
away from the shelves, which could have eased the motion. One possibility to extend the
robot accessibility space is to automatically change the support polygon to solve the task.
This is an ongoing research field, \cite{OussamaKanoun06102010} proposes some 
local solutions. This has not been integrated in our global planning algorithm yet.

@@

A  robotic system  is  controllable  if  for any two  configurations
$q_1$ and $q_2$,  there exists  a
trajectory  going  from  $q_1$ to  $q_2$.  It  is  
\textit{small-space  controllable} if for  any configuration  $q$, 
for any $\epsilon >0$, there
exist $\eta >0$ such that all the configurations contained in the ball of center
$q$ and radius $\eta$ are reachable by trajectories included in the
ball of center $q$ and radius $\epsilon$. Fig.~\ref{fig:ssc1} shows an illustration
of this property.

\begin{figure}[h]
  \centering
  \input{stc1.tex}

  \caption{The small-space controllability local property:  any configuration $q'$ 
    at distance less than
    $\eta$ is reachable from $q$ by an admissible trajectory included in
    a ball of size $\epsilon$.}
  \label{fig:ssc1}
\end{figure}

@@

Fig. \ref{fig:ssc2} shows an example of collision-free 
path approximation by admissible collision-free sub-trajectories. The fact 
that this algorithm  converges is guaranteed by the small-space
controllability property.

@@

  \input{stc2.tex}

  \caption{Small-space controllability in motion planning. 
    A collision-free path from
    $q_1$ to $q_2$ is approximated by collision-free and admissible
    trajectories by using the local property.
  }
  \label{fig:ssc2}
\end{figure}

@@

\begin{proof}[Proof of Theorem \ref{thm:ssc}]
  Let $\mathcal{C}$ be the configuration space of a small-space controllable robot, and 
  $\mathcal{C}_{free} \subset \mathcal{C}$ the set of collision-free configurations. We
  consider in-contact configurations as colliding, so $\mathcal{C}_{free}$ is an open set.
  Let $\tau : [0,1] \rightarrow \mathcal{C}_{free}$ be a collision-free path. For every $x \in [0,1]$,
  $\tau(x) \in \mathcal{C}_{free}$, so there exists $\epsilon_x$ such that the open ball 
  $B(\tau(x),\epsilon_x)$ of center $\tau(x)$ and radius $\epsilon_x$ is included in 
  $\mathcal{C}_{free}$. The small-space controllability property states that for every $x$,
  there exist $\eta_x > 0$ such that every configuration $q \in B(\tau(x),\eta_x)$ is reachable 
  from $\tau(x)$ by a trajectory included in $B(\tau(x),\epsilon_x)$.

@@

\begin{theorem}
\label{thm:humanoid-ssc}
A quasi-static walking robot is not small-space controllable. A dynamic walking robot is.
\end{theorem}

\begin{proof}[Proof of Theorem \ref{thm:humanoid-ssc}]

The first claim is straightforward. Let the robot be in a configuration $q$ where the two
feet are separated from a positive distance. Let  $L>0$ be the positive horizontal distance
between the CoM and the left foot (if the CoM is over the left foot, we can consider similarly 
the right foot). For any $\epsilon < L$, any valid trajectory starting from $q$, included in
the ball of center $q$ and radius $\epsilon$, is such that the CoM is never over the left foot.
Given the quasi-static walk conditions, the right foot of the robot is fixed along 
the trajectory. Thus, the set of accessible configurations from $q$ by staying inside 
$B(q,\epsilon)$ does not form a neighborhood of $q$, since it does not include any configuration
where the right foot has moved. This shows that the robot is not small-space controllable.

@@

Without loss of generality, let us assume that the robot is in a configuration
in which the CoM is at the horizontal position $(0,0)$, the center of the feet are 
aligned with the $y$-axis and the horizontal distance between the CoM and either of the centers
of the feet is $L$. To achieve dynamically balanced walking, we aim at making
$p_y(t)$ oscillate  between $-L$ and $L$. To move the ZMP  under a
given foot, only  the $y$ coordinate of the CoM  is of interest. Thus,
we will keep the $x$ coordinates  of the CoM and ZMP constant equal to
$0$. By hypothesis, the feet are of positive surface,
let $l>0$ be such that the length of the section of a foot along the 
$y$-axis is greater than $l$. Fig.~\ref{fig:simple-humanoid} summarizes the notations used
in the following.

\begin{figure}[h]
  \centering
  \input{humanoid.tex}

  \caption{Simplified model of a legged robot. The CoM is at $(0,0,z_c)$, the two feet 
    are flat on the ground, aligned with the $y$-axis, at a horizontal distance $L$ 
    from the CoM.}
  \label{fig:simple-humanoid}
\end{figure}

The idea of this proof is to use the form of Eq. (\ref{eq:walk-zmp}) to
apply a  scaling factor between  the amplitude of the  oscillations of
the CoM and of the ZMP. For $\omega >0$, assuming the CoM follows the trajectory
$y(t) = \epsilon \sin(\omega t)$,  Eq. (\ref{eq:walk-zmp}) gives:
\[
p_y(t) =
(1+\left(\frac{\omega}{\omega_0}\right)^2)\epsilon\sin(\omega t)
\]

The
amplitude  of  the oscillations  of  $y$  is  multiplied by  a  factor
$(1+\left(\frac{\omega}{\omega_0}\right)^2)$.  Choosing  $\omega =
\omega_0 \sqrt{\frac{L}{\epsilon} -1}$ makes  $p_y$ oscillates between $-L$
and    $L$, while $y$ oscillate between $-\epsilon$ and $\epsilon$.   
At    time   $t_l^{(n)}    =    n\frac{2\pi}{\omega}   +
\frac{\pi/2}{\omega}$, the  ZMP is located  at the center of  the left
foot,  the robot  can move  its right  foot and  at time  $t_r^{(n)} =
n\frac{2\pi}{\omega}  + \frac{3\pi/2}{\omega}$ the  ZMP is  located at
the center of the right foot, the robot can move its left foot.

@@

One can  check that $y$  is of class $C^2$ over  $\mathbb{R}_+$, and
that $\dot{f}(0) = 0$. When $t\geq T$, the robot is in the permanent
state described above  and can successively move either of its 
feet  inside small neighborhoods.
The  last point to check
is that for $t \in  [0,T]$ $p_y(t)$ stays inside the support polygon
of the robot. The calculation of the successive derivatives of $y$ gives:

@@

We have thus exhibited a continuous control scheme that allows to move any of the feet
in any direction, while keeping the CoM inside an arbitrarily small neighborhood. This concludes
the proof.
\end{proof}

\subsubsection{Remarks}
\paragraph{Generalization to a complete model:} More work is needed to generalize 
the previous proof to any legged robot model. First, because we use a model where
the mass is punctual, we did not have to consider the angular-momentum of the robot
around the CoM, while it accounts for dynamic balance. Moreover, when considering
an actual kinematic and dynamic model,
the positions of feet and CoM are not controlled degrees of freedom, but inputs 
for inverse kinematic tasks. Continuity and regularity results about the robot
controller -- i.e. how much the configuration changes when the input of an inverse
kinematics task changes--, would be needed to generalize the proof. We have not
proven such properties for the inverse kinematics controller we use. 
The  proof presented above still gives the intuition for the
algorithm presented in the following.

\paragraph{Use of ZMP preview controller:} The control strategy presented in the  
previous proof may
generate very  long trajectories, because  of the transient  states at
the beginning and end of the locomotion. In the actual implementation,
we have chosen  to generate  CoM motions with  a ZMP preview  controller, as
presented in \cite{kajita2003biped}.  We  have observed experimentally that the
amplitude of  CoM trajectories decreases  when the frequency  of steps
increases. Our current ZMP preview controller relies on the cart-table simplified model.
To make this approximation valid, we fix the height of the robot CoM during walk,
as well as the vertical orientation of the robot waist. These geometric constraints 
are also applied when planning a statically stable path. Note that this is
due to our current implementation of ZMP preview controller, and does not affect
the generality of the small-space controllability result presented above.

\paragraph{Speed of CoM:} The theoretical result presented in this section implies
that any collision-free path can be approximated by a sequence of admissible 
and collision-free trajectories. However, the theorem depends on a control law
that generates trajectories with unbounded velocities for the CoM, when the input
path is close to obstacles. The humanoid robot
hardware may be a limitation to such trajectories. To prevent the generated CoM 
oscillations from being too fast, one has to require that the statically stable 
path is included
inside an $\epsilon$-radius tube of the free space, where $\epsilon$ depends on the
physical capacities of the robot.

@@

The  algorithm   that  animates  a  statically  stable   path  into  a
dynamically stable  walk trajectory has been inspired  by the previous
small-space controllability  proof.
Given a statically  stable path $p$,
we  start  by placing  footsteps  corresponding  to  the nominal  walk
pattern  of the  robot. Given the footsteps, we  compute a ZMP trajectory, 
and a preview controller outputs  a corresponding CoM trajectory, using
the cart-table model. The stack of tasks applied to
the robot to generate a dynamic walk motion is -- in decreasing priority order:

@@

\item Positions and orientations of  feet,
